---
title: "Global Macro"
output: pdf_document 
date: "2025-05-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# US Macro Regimes on Global Markets

## Preparation

### Libraries

```{r include=FALSE}

if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  quantmod,
  dplyr,
  tidyquant,
  TTR,
  tidyr,
  corrplot,
  MASS,
  Rblpapi,
  tidyverse,
  tibble,
  pheatmap,
  quadprog,
  frenchdata,
  lubridate,
  purrr,
  broom,
  kableExtra
)


```


### Get Necessary Data

```{r}
load("Minimal_Data.RData")
```


1. US Macroeconomic Data

```{r}
start_date_macro <- as.Date("1988-01-01")

# Download macro data from FRED --> GDP, Consumer Price Index, Fed Funds Rate
getSymbols(c("GDP", "CPIAUCSL", "FEDFUNDS"), src = "FRED", from = start_date_macro)

# Load NBER recession dates
nber_dates <- tq_get("USREC", get = "economic.data", from = start_date_macro)
nber_xts <- xts(nber_dates[["price"]], order.by = nber_dates$date)

# Merge data (monthly basis)
macro_data <- merge(
  GDP,
  CPIAUCSL,
  FEDFUNDS,
  nber_xts
)
colnames(macro_data) <- c("GDP", "CPI", "FedFundsRate","NBER")

macro_data_df <- data.frame(date = index(macro_data), coredata(macro_data))

# Apply lags for information availability
macro_data_df <- macro_data_df %>%
  mutate(
    GDP = lag(GDP, 2),
    CPI = lag(CPI, 1),
  )

# Forward-fill for monthly completeness
macro_data_df[ , 2:5] <- na.locf(macro_data_df[ , 2:5], na.rm = FALSE)

# Caluclate YoY Growth
macro_data_df <- macro_data_df %>%
  mutate(
    GDP_YoY = ROC(GDP, n = 12, type = "discrete"),
    CPI_YoY = ROC(CPI, n = 12, type = "discrete")
  )


# Create Regimes

## GDP

macro_data_df <- macro_data_df %>%
  mutate(growth_regime = if_else( NBER == 1, "Low", "High"))

## Inflation

macro_data_df <- macro_data_df %>%
  mutate(
    CPI_YoY_lag = lag(CPI_YoY, 12),
    inflation_surprise = CPI_YoY - CPI_YoY_lag,
    inflation_regime = case_when(
      CPI_YoY > 0.02 & inflation_surprise > 0 ~ "High",
      TRUE ~ "Low"
    )
  )

## Fed Rate

macro_data_df <- macro_data_df %>%
  mutate(
    FedChange = FedFundsRate - lag(FedFundsRate),
    FedRegime = case_when(
      FedChange > 0 ~ "Hiking",
      FedChange < 0 ~ "Cutting",
      TRUE ~ "Flat"
    )
  )

## Classic Regimes

macro_data_df <- macro_data_df %>%
  mutate(
    macro_regime = case_when(
      growth_regime == "High" & inflation_regime == "Low" ~ "Goldilocks",
      growth_regime == "High" & inflation_regime == "High" ~ "Overheating",
      growth_regime == "Low"  & inflation_regime == "Low" ~ "Downturn",
      growth_regime == "Low"  & inflation_regime == "High" ~ "Stagflation"
    )
  )

## Check Fed Rate

macro_data_df <- macro_data_df %>%
  mutate(
    combined_regime = paste(macro_regime, FedRegime, sep = " + ")
  )
macro_data_df <- na.omit(macro_data_df)

# Check Regimes
Regimes <- unique(macro_data_df$macro_regime)


```

2. Country ETF


```{r}
##### Bloomberg Data ######
# blpConnect()
# 
# securities <- c(
#                 ### Emerging Markets ###
#                 "MSEUSIA Index",  # India
#                 "MSEUSCF Index",  # China
#                 "MSIUEG Index",   # Egypt
#                 "MSEUSSA Index",  # South Africa
#                 "MSEUSKO Index",  # South Korea
#                 "MSEUSTW Index",  # Taiwan
#                 "MSEUSTHF Index", # Thailand
#                 "MSEUSCZ Index",  # Czeck Republic
#                 "MSEUSPO Index",  # Poland
#                 "MXAR Index",     # Argentina
#                 "MXCL Index",     # Chile
#                 "M3GR Index",     # Greeece
#                 "MSEUSTK Index",  # Turkey
# 
#                 ### Developed Market ###
#                 "MSDUCA Index",   # Canada
#                 "MXUS Index",     # USA
#                 "MSDUFR Index",   # France
#                 "MSDUGR Index",   # Germany
#                 "MXIL Index",     # Israel
#                 "MSDUIT Index",   # Italy
#                 "MSDUSW Index",   # Sweden
#                 "MSDUSP Index",   # Spain
#                 "MSDUSZ Index",   # Switzerland
#                 "MSDUUK Index",   # United Kingdom
#                 "MSDUAS Index",   # Australia
#                 "MSDUHK Index",   # Hong Kong
#                 "MSDUJN Index",   # Japan
#                 "MSDUSG Index"    # Singapore
# )
# 
# countries <- c(
#   ### Emerging Markets ###
#   "India",
#   "China",
#   "Egypt",
#   "South Africa",
#   "South Korea",
#   "Taiwan",
#   "Thailand",
#   "Czech Republic",
#   "Poland",
#   "Argentina",
#   "Chile",
#   "Greece",
#   "Turkey",
# 
#   ### Developed Market ###
#   "Canada",
#   "USA",
#   "France",
#   "Germany",
#   "Israel",
#   "Italy",
#   "Sweden",
#   "Spain",
#   "Switzerland",
#   "United Kingdom",
#   "Australia",
#   "Hong Kong",
#   "Japan",
#   "Singapore"
# )
# 
# 
# start_date <- as.Date("1990-01-01")
# 
# end_date <- Sys.Date()
# 
# Fields <- c("PX_LAST")
# 
# 
# etf_close_df <- data.frame()
# 
# for(i in 1:length(securities)) {
#   # Get Bloomberg data
#   security_df <- bdh(securities[i], fields = "PX_LAST", start.date = start_date, end.date = end_date)
# 
#   # Rename the PX_LAST column to the ticker name
#   colnames(security_df)[2] <- countries[i]
# 
#   # Merge using full_join on the date column
#   if (nrow(etf_close_df) == 0) {
#     etf_close_df <- security_df
#   } else {
#     etf_close_df <- full_join(etf_close_df, security_df, by = "date")
#   }
# }
# colnames(etf_close_df) <- c("date", countries)
# 
# 
# etf_close_xts <- xts(etf_close_df[,-1], order.by = etf_close_df$date)

# Calculate Returns
etf_close_xts_weekly <- apply.weekly(etf_close_xts, first)
etf_returns_xts <- ROC(etf_close_xts_weekly, n = 1, type = "discrete")

# Create dataframe with dates labeled as start-of-week
etf_returns_df <- data.frame(
  date = index(etf_returns_xts),
  coredata(etf_returns_xts)
)

# Remove NA at the beginning (from ROC calculation)
etf_returns_df <- etf_returns_df[-1, ]

```

# Inlude Trade Deficit to the US

## Get Exporting Data to the US 

```{r}
# # Reporting country codes from IMF DOTS (to US = 111)
# reporting_codes <- c(
#   "India"           = "534",
#   "China"           = "924",
#   "Egypt"           = "469",
#   "South.Africa"    = "199",
#   "South.Korea"     = "542",
#   "Taiwan"          = "528",
#   "Thailand"        = "578",
#   "Czech.Republic"  = "939",
#   "Poland"          = "964",
#   "Argentina"       = "213",
#   "Chile"           = "228",
#   "Greece"          = "174",
#   "Turkey"          = "186",
#   "Canada"          = "156",
#   "USA"             = "111",
#   "France"          = "132",
#   "Germany"         = "134",
#   "Israel"          = "436",
#   "Italy"           = "136",
#   "Sweden"          = "144",
#   "Spain"           = "184",
#   "Switzerland"     = "146",
#   "United.Kingdom"  = "112",
#   "Australia"       = "193",
#   "Hong.Kong"       = "532",
#   "Japan"           = "158",
#   "Singapore"       = "576"
# )
# 
# length(reporting_codes)
# 
# # Build the Bloomberg tickers for Exports to the U.S.
# export_tickers <- paste0(reporting_codes, "D0111 Index")
# names(export_tickers) <- names(reporting_codes)
# 
# # View the tickers
# export_tickers
# 
# 
# get_trading_data <- function(ticker) {
#   tryCatch({
#     bdh(ticker, "PX_LAST", start.date = as.Date("1190-02-01"))
#   }, error = function(e) NULL)
# }
# 
# # Loop over all tickers
# export_data_list <- lapply(export_tickers, get_trading_data)
# names(export_data_list) <- names(export_tickers)
# 
# export_df <- bind_rows(export_data_list, .id = "country")
# 
# 
# length(unique(export_df$country))
# 
# 
# colnames(export_df)[3] <- "Exports_US"
```

## Get Trade Imports to the US

```{r}
# import_tickers <- paste0("111D0", reporting_codes, " Index")
# names(import_tickers) <- names(reporting_codes)
# 
# 
# 
# # Download import data for each country
# import_data_list <- lapply(import_tickers, get_trading_data)
# names(import_data_list) <- names(import_tickers)
# 
# import_df <- bind_rows(import_data_list, .id = "country")
# 
# colnames(import_df)[3] <- "Imports_US"
# 
# Trade_data <- left_join(export_df, import_df, by = c("date", "country"))

```

## Calcualte the Trade Deficit/Surplus

```{r}
Trade_data <- Trade_data %>%
  mutate(
    Trade_Balance = Exports_US - Imports_US
  )

Trade_data_summary <- Trade_data %>%
  group_by(country)%>%
  summarize(
    Mean = mean(Trade_Balance),
    variability = sd(Trade_Balance)
  )

Trade_data_summary
```


## Create weekly base for the performance analysis


```{r}

## Combine the weekly returns with the monthly macroeconomic data to create a bigger data base for the optimization

## Create a new column with the corresponding month of the weekly return which is used to merge the data


etf_returns_df <- etf_returns_df %>%
  mutate(month_date = floor_date(date, "month"))

## Merge data according to the month date column

# Create long format dataframe before merging
etf_returns_long <- etf_returns_df %>%
  pivot_longer(-c(date, month_date), names_to = "ETF", values_to = "return")

# Merge the data
returns_with_regimes <- etf_returns_long %>%
  left_join(
    macro_data_df %>%
      dplyr::select(date, macro_regime,combined_regime) %>%
      rename(month_date = date),
    by = "month_date") %>%
  dplyr::select(-month_date)


returns_with_regimes <- returns_with_regimes %>%
  filter(date >= as.Date("1990-02-01")) %>%
  filter(!is.na(macro_regime))
  
knitr:: kable(returns_with_regimes %>%
  group_by(macro_regime) %>%
  summarise(n_obs = sum(!is.na(return))),
  caption = "Observations in each regime" )

```

## Helper for missing data 

```{r}
remove_na_heavy <- function(df, row_thresh = 0.5, col_thresh = 0.5,
                            interpolate = FALSE) {
  if (!is.data.frame(df)) {
    stop("Input must be a data frame.")
  }
  
  # Convert percentage to proportion if needed
  if (row_thresh > 1) row_thresh <- row_thresh / 100
  if (col_thresh > 1) col_thresh <- col_thresh / 100
  
  # Step 1: Drop rows with too many NAs
  row_na_prop <- rowMeans(is.na(df))
  df <- df[row_na_prop <= row_thresh, , drop = FALSE]
  
  # Step 2: Drop columns with too many NAs
  col_na_prop <- colMeans(is.na(df))
  df <- df[, col_na_prop <= col_thresh, drop = FALSE]
  
  # Step 3: Interpolate missing values (optional)
  if (interpolate) {
    is_numeric <- sapply(df, is.numeric)
    
    df[is_numeric] <- lapply(df[is_numeric], function(column) {
      na_index <- is.na(column)
      if (all(na_index)) return(column)  # skip all-NA columns
      approx(x = which(!na_index),
             y = column[!na_index],
             xout = seq_along(column),
             rule = 2)$y  # rule = 2 → use nearest value for extrapolation
    })
  }
  
  return(df)
}

```

## Performance Analysis

```{r}
summarise_regime <- function(data, regime_name) {
  
  performance_list <- list()
  correlation_list <- list()
  portfolio_list <- list()
  
  Port_Ret <- numeric()
  Port_Sd <- numeric()
  Port_Sharpe <- numeric()
  
  data <- as_tibble(data)
  
  for(regime in regime_name){
  df <- data %>%
    filter(macro_regime == regime) %>%
    dplyr::select(date, ETF, return) %>%
    pivot_wider(names_from = ETF, values_from = return)
  
   # Keep only ETFs with at least some data
  valid_cols <- colSums(!is.na(df[ , -1])) > 0
  df <- df[ , c(TRUE, valid_cols)]  
  df <- df[rowSums(is.na(df[ , -1])) < ncol(df) - 1, ]
  
  #Annualized stats
  returns_mat <- as.matrix(df[ , -1])
  returns_mat <- returns_mat[ , colSums(!is.na(returns_mat)) >= 52]
  mean_returns <- colMeans(returns_mat, na.rm = TRUE) * 52
  US_mean <- mean(returns_mat[,"USA"],na.rm = TRUE) * 52
  mean_excess_return_US <- mean_returns - US_mean 
  sd_returns <- apply(returns_mat, 2, sd,na.rm = TRUE) * sqrt(52)
  sharpe_ratios <- mean_returns / sd_returns
  
  # Assemble summary table
  perf_df <- data.frame(
    ETF = colnames(returns_mat),
    AnnualReturn = round(mean_returns, 4),
    AnnualVolatility = round(sd_returns, 4),
    AnnualExcessRetUS = round(mean_excess_return_US,4),
    Sharpe = round(sharpe_ratios, 2)
  ) %>%
    arrange(desc(Sharpe))
  
  performance_list[[regime]] <- perf_df
  
  ### Correlation ###
  corr_matrix <- round(cor(returns_mat, use = "pairwise.complete.obs"), 2)
  
  correlation_list[[regime]] <- corr_matrix
  
 ### Mean-Variance Portfolio Optimization ###
  
  # Expectation and Covariance Matrix
  returns_mat <- as.matrix(remove_na_heavy(as.data.frame(returns_mat), row_thresh = 0.5,
                                           col_thresh = 0.3,interpolate = TRUE))
  
  mu <- colMeans(returns_mat, na.rm = TRUE) * 52
  Sigma <- cov(returns_mat, use = "pairwise.complete.obs") * 52
  # Dmat and dvec for solve.QP
  Dmat <- 2 * Sigma  # QP minimizes (1/2) * w' D w, so we multiply by 2
  dvec <- rep(0, length(mu))
  
  # Constraints: sum(weights) = 1, and weights >= 0
  n_assets <- length(mu)
  # Equality constraint: sum(weights) = 1
  A_eq <- matrix(1, nrow = n_assets, ncol = 1)
  # Inequality constraints: weights >= 0
  A_ineq <- diag(n_assets)
  # Combine constraint matrix
  Amat <- cbind(A_eq, A_ineq)  # Each column is a constraint
  # Right-hand side of constraints
  bvec <- c(1, rep(0, n_assets))  # sum(weights) = 1 and weights >= 0
  # Solve QP
  sol <- solve.QP(Dmat, dvec, Amat, bvec, meq = 1)
  
  weights <- sol$solution
  
  port_ret <- as.numeric(t(weights) %*% mu)
  port_var <- as.numeric(t(weights) %*% Sigma %*% weights)
  port_sd <- sqrt(pmax(port_var, 0))
  port_sharpe <- port_ret / port_sd

  mean_var_df <- data.frame(
    ETF = colnames(returns_mat),
    Weights = round(as.numeric(weights), 6)
  )
  
  Port_Ret[regime] <- port_ret
  Port_Sd[regime] <- port_sd
  Port_Sharpe[regime] <- port_sharpe
  portfolio_list[[regime]] <- mean_var_df
  
  }

  port_per_df <- data.frame(
    Regime = names(Port_Ret),
    Portfolio_Return = Port_Ret,
    Portfolio_SD = Port_Sd,
    Portfolio_Sharpe = Port_Sharpe
  )


  return(list(
    performance = performance_list,
    correlation = correlation_list,
    portfolio =  portfolio_list,
    portfolio_perf = port_per_df
  ))
}

results <- summarise_regime(returns_with_regimes, Regimes)
```

## Plot Correlations

```{r, fig.width=8, fig.height=8}
for (regime in Regimes) {
  corr_matrix <- results$correlation[[regime]]
  
  corrplot(
    corr_matrix,
    method = "color",
    type = "upper",
    order = "hclust",
    addCoef.col = "black",
    tl.cex = 0.7,         # Smaller label size
    number.cex = 0.6,     # Smaller number size
    col = colorRampPalette(c("blue", "white", "red"))(200),
    cl.pos = "n",
    title = paste("Correlation –", regime),
     mar = c(1, 1, 3, 1),
  )
}

```

## Table Outputs 

```{r, results = 'asis'}
for (regime in Regimes) {
  cat("### Performance in", regime, "\n\n")
  
  performance_df <- results$performance[[regime]]
  colnames(performance_df)[1] <- "country"
  
  output <- performance_df %>%
  left_join(Trade_data_summary %>% dplyr::select(country, Mean), by = "country") %>%
    rename(Avg.Trade.Balance = Mean) %>%
    mutate(Avg.Trade.Balance = round(Avg.Trade.Balance,2)) %>%
  mutate(Avg.Trade.Balance = ifelse(is.na(Avg.Trade.Balance), "/", Avg.Trade.Balance))
  
  
  rownames(output) <- NULL
  
  print(knitr::kable(
    output,
    digits = 3,
    caption = paste("Country Performance in", regime, "regime")
  ))
  # Page break after each table (except the last)
    if(regime != tail(Regimes, 1)) {
    cat("\\newpage\n\n")
    }
  
}

```

## Plot Correlation to USA

```{r}
us_cor_df <- lapply(Regimes, function(regime) {
  mat <- results$correlation[[regime]]
  if (is.null(mat)) return(NULL)
  cor_us <- mat[, "USA"]
  data.frame(
    ETF = names(cor_us),
    Correlation = cor_us,
    Regime = regime
  )
}) %>% bind_rows()

pheatmap(
  pivot_wider(us_cor_df, names_from = Regime, values_from = Correlation) %>%
    column_to_rownames("ETF"),
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  main = "U.S. Correlation by Regime"
)
```
## Plot Portfolio Weights

```{r}
portfolio_list_renamed <- imap(results$portfolio, ~ rename(.x, !!.y := Weights))
portfolio_merged <- reduce(portfolio_list_renamed, full_join, by = "ETF")

## Bar Chart
portfolio_merged_long <- pivot_longer(portfolio_merged,cols = -ETF, names_to = "Portfolio", values_to = "Weight")
ggplot(portfolio_merged_long, aes(x = ETF, y = Weight, fill = Portfolio)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(title = "Portfolio Weights by Country", y = "Weight", x = "Asset Class") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    plot.title = element_text(hjust = 0.5)
  ) +
  coord_flip()
```

## Plot Portfolio Performance

```{r}
Performance_Analysis <- results$portfolio_perf
rownames(Performance_Analysis) <- NULL
knitr::kable(cbind(Regimes,Performance_Analysis), digits = 3,
             caption = "Performance Analysis MVO Portfolios")
```



## Explanatory Regression

### Merge Everything and include explanatory variables on monthly basis

```{r include=FALSE}

## Calculate monthly returns
etf_close_xts_monthly <- apply.monthly(etf_close_xts, first)
etf_returns_xts_monthly <- ROC(etf_close_xts_monthly, n = 1, type = "discrete")

# Create dataframe with dates labeled as start-of-week
etf_returns_monthly_df <- data.frame(
  date = as.Date(as.yearmon(index(etf_returns_xts_monthly)),frac = 0),
  coredata(etf_returns_xts_monthly)
)

# Remove NA at the beginning (from ROC calculation)
etf_returns_monthly_df <- etf_returns_monthly_df[-1, ]

etf_returns_monthly_df < etf_returns_monthly_df[-nrow(etf_returns_monthly_df)]

etf_returns_long_monthly <- etf_returns_monthly_df %>% 
  pivot_longer(-date,names_to = "country", values_to = "return")

# Combine with the regimes

returns_with_regimes_monthly <- etf_returns_long_monthly %>%
  left_join(macro_data_df %>% dplyr::select(date,macro_regime,combined_regime, inflation_surprise, FedChange, growth_regime), by = "date")

# Combine with the trade Balance

Trade_data <- Trade_data %>%
  mutate(
    date = as.Date(as.yearmon(date),frac = 0),
    trade_change = (Trade_Balance - lag(Trade_Balance)) / Trade_Balance
  )

returns_with_regimes_monthly <- returns_with_regimes_monthly  %>%
  left_join(Trade_data %>% dplyr::select(date,country, trade_change, Trade_Balance), by = c("country","date")) %>%
  filter(!country %in% c("USA", "Taiwan"),!is.na(macro_regime))




```



```{r}
regression_results <- returns_with_regimes_monthly %>%
  group_by(country) %>%
  nest() %>%
  mutate(
    # Run linear model per country
    model = map(data, ~ lm(return ~ inflation_surprise + FedChange + growth_regime + trade_change, data = .x)),
    # Extract coefficients with heteroskedasticity-robust SEs
    tidy_out = map(model, ~ tidy(.x, se.type = "HC1")),
    # Extract model-level stats (R², F-statistic, etc.)
    glance_out = map(model, glance)
  )

# Unnest tidy results and keep estimate + p-value
coefficients_df <- regression_results %>%
  dplyr::select(country, tidy_out) %>%
  unnest(tidy_out) %>%
  dplyr::select(country, term, estimate, p.value)

# Pivot estimate and p-value into long format
coefficients_long <- coefficients_df %>%
  pivot_longer(cols = c(estimate, p.value),
               names_to = "stat_type",
               values_to = "value") %>%
  mutate(stat_name = paste0(term, "_", stat_type)) %>%
  dplyr::select(country, stat_name, value)

# Pivot to wide format: one row per country, estimates and p-values side-by-side
coefs_wide <- coefficients_long %>%
  pivot_wider(names_from = stat_name, values_from = value)

model_stats_df <- regression_results %>%
  dplyr::select(country, glance_out) %>%
  unnest(glance_out) %>%
  dplyr::select(
    country,
    r.squared,
    adj.r.squared,
    f_pvalue = p.value
  )

output <- coefs_wide %>%
  left_join(model_stats_df, by = "country")

output <- output %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

colnames(output) <- c("Country","Inter.", "p-V.", "Infl. Surp.","p-V.", "Fed Change","p-V.", "L.Growth","p-V.", "TB Change","p-V.", "R^2","F-T.","p-V.")

# Step 7: Table output
knitr::kable(output, caption = "Explanatory Regression by Country") %>%
  kable_styling(latex_options = c("scale_down"), font_size = 6)

```

## Split the regression into multiple Regimes

```{r}
# Step 1: Fit model per country and regime
regression_results <- returns_with_regimes_monthly %>%
  group_by(country, macro_regime) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(return ~ inflation_surprise + FedChange + Trade_Balance, data = .x)),
    tidy_out = map(model, ~ tidy(.x, se.type = "HC1")),
    glance_out = map(model, glance)
  )

# Step 2: Extract tidy (coefficients) and pivot wide
tidy_df <- regression_results %>%
  dplyr::select(country, macro_regime, tidy_out) %>%
  unnest(tidy_out) %>%
  dplyr::select(country, macro_regime, term, estimate, p.value)

coefs_df <- tidy_df %>%
  pivot_wider(
    names_from = term,
    values_from = c(estimate, p.value),
    names_glue = "{term}_{.value}"
  )

# Step 3: Extract model-level stats
glance_df <- regression_results %>%
  dplyr::select(country, macro_regime, glance_out) %>%
  unnest(glance_out) %>%
  dplyr::select(country, macro_regime, r.squared, adj.r.squared, statistic, p.value) %>%
  rename(
    R_squared = r.squared,
    Adj_R_squared = adj.r.squared,
    F_statistic = statistic,
    F_p_value = p.value
  )

# Step 4: Join and format output
output <- coefs_df %>%
  left_join(glance_df, by = c("country", "macro_regime")) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  arrange(country, macro_regime)

# Step 5: Output as table (optional: use kableExtra for better formatting)
knitr::kable(output, caption = "Explanatory Regressions by Country and Regime") %>%
kable_styling(latex_options = c("scale_down"), font_size = 8)
```
#################################################################################
# Fama French Factors
#################################################################################

### Download the data from Fama French

```{r}
start_date_famafrench <- as.Date("1954-08-01")

start_date_famafrench <- as.Date("1954-08-01")

################# Size - Book to Market ######################################

# 1. Download daily Fama-French portfolios
ff3_ports_raw_sb <- download_french_data("6 Portfolios Formed on Size and Book-to-Market (2 x 3) [Daily]")
ff3_tbl_sb <- ff3_ports_raw_sb$subsets$data[[1]]

# 2. Clean and transform
colnames(ff3_tbl_sb) <- c("date", "Small_Low.B/M", "Small_Mid.B/M", "Small_High.B/M",
                             "Big_Low.B/M", "Big_Mid.B/M", "Big_High.B/M")

ff_ports_daily_sb <- ff3_tbl_sb %>%
  # Convert date from integer YYYYMMDD to proper Date format
  mutate(date = ymd(as.character(date))) %>%
  # Convert returns from percent to decimal
  mutate(across(-date, ~ as.numeric(.) / 100)) %>%
  # Rename columns to lowercase for consistency
  rename_with(str_to_lower) %>%
  # Filter by start date
  filter(date >= start_date_famafrench)

# 3. Convert to xts object
ff_xts_sb <- xts(ff_ports_daily_sb %>% dplyr::select(-date), order.by = ff_ports_daily_sb$date)

# 4. Convert to weekly returns (simple or log)
ff_weekly_sb <- apply.weekly(ff_xts_sb, Return.cumulative)  # Or use Return.calculate for log returns

# Create dataframe with dates labeled as start-of-week
ff_size_book <- data.frame(
  date = floor_date(index(ff_weekly_sb), unit = "week", week_start = 1),
  coredata(ff_weekly_sb)
)

########################## Size Momentum ######################################


# 1. Download daily Fama-French portfolios
ff3_ports_raw_sm <- download_french_data("6 Portfolios Formed on Size and Momentum (2 x 3) [Daily]")
ff3_tbl_sm <- ff3_ports_raw_sm$subsets$data[[1]]

# 2. Clean and transform
colnames(ff3_tbl_sm) <- c("date", "Small_Low.Mom", "Small_Mid.Mom", "Small_High.Mom",
                             "Big_Low.Mom", "Big_Mid.Mom", "Big_High.Mom")

ff_ports_daily_sm <- ff3_tbl_sm %>%
  # Convert date from integer YYYYMMDD to proper Date format
  mutate(date = ymd(as.character(date))) %>%
  # Convert returns from percent to decimal
  mutate(across(-date, ~ as.numeric(.) / 100)) %>%
  # Rename columns to lowercase for consistency
  rename_with(str_to_lower) %>%
  # Filter by start date
  filter(date >= start_date_famafrench)

# 3. Convert to xts object
ff_xts_sm <- xts(ff_ports_daily_sm %>% dplyr::select(-date), order.by = ff_ports_daily_sm$date)

# 4. Convert to weekly returns (simple or log)
ff_weekly_sm <- apply.weekly(ff_xts_sm, Return.cumulative)  # Or use Return.calculate for log returns

# Create dataframe with dates labeled as start-of-week
ff_size_mom <- data.frame(
  date = floor_date(index(ff_weekly_sm), unit = "week", week_start = 1),
  coredata(ff_weekly_sm)
)

ff_portfolios <- ff_size_mom %>%
  left_join(ff_size_book, by = "date")

```

### Combine

```{r}

ff_portfolios <- ff_portfolios %>%
  mutate(month_date = floor_date(date, "month"))


ff_ports_long <- ff_portfolios %>%
  pivot_longer(-c(date, month_date), names_to = "Portfolio", values_to = "return")

# Merge the data
ff_returns_with_regimes <- ff_ports_long %>%
  left_join(
    macro_data_df %>%
      dplyr::select(date, macro_regime,combined_regime) %>%
      rename(month_date = date),
    by = "month_date") %>%
  dplyr::select(-month_date)


ff_returns_with_regimes <- ff_returns_with_regimes %>%
  filter(date >= as.Date("1990-02-01")) %>%
  filter(!is.na(macro_regime))
 

knitr:: kable(ff_returns_with_regimes %>%
  group_by(macro_regime) %>%
  summarise(n_obs = sum(!is.na(return))))

```

## Portfolio Analysis

### Portfolio

```{r}
summarise_regime <- function(data, regime_name) {
  
  performance_list <- list()
  correlation_list <- list()
  portfolio_list <- list()
  
  Port_Ret <- numeric()
  Port_Sd <- numeric()
  Port_Sharpe <- numeric()
  
  data <- as_tibble(data)
  
  for(regime in regime_name){
  df <- data %>%
    filter(macro_regime == regime) %>%
    dplyr::select(date, Portfolio, return) %>%
    pivot_wider(names_from = Portfolio, values_from = return)
  
   # Keep only ETFs with at least some data
  valid_cols <- colSums(!is.na(df[ , -1])) > 0
  df <- df[ , c(TRUE, valid_cols)]  
  df <- df[rowSums(is.na(df[ , -1])) < ncol(df) - 1, ]
  
  #Annualized stats
  returns_mat <- as.matrix(df[ , -1])
  returns_mat <- returns_mat[ , colSums(!is.na(returns_mat)) >= 52]
  mean_returns <- colMeans(returns_mat, na.rm = TRUE) * 52
  sd_returns <- apply(returns_mat, 2, sd,na.rm = TRUE) * sqrt(52)
  sharpe_ratios <- mean_returns / sd_returns
  
  # Assemble summary table
  perf_df <- data.frame(
    Portfolios = colnames(returns_mat),
    AnnualReturn = round(mean_returns, 4),
    AnnualVolatility = round(sd_returns, 4),
    Sharpe = round(sharpe_ratios, 2)
  ) %>%
    arrange(desc(Sharpe))
  
  performance_list[[regime]] <- perf_df
  
  ### Correlation ###
  corr_matrix <- round(cor(returns_mat, use = "pairwise.complete.obs"), 2)
  
  correlation_list[[regime]] <- corr_matrix
  
 ### Mean-Variance Portfolio Optimization ###
  
  # Expectation and Covariance Matrix
  returns_mat <- as.matrix(remove_na_heavy(as.data.frame(returns_mat), row_thresh = 0.5,
                                           col_thresh = 0.3,interpolate = TRUE))
  
  mu <- colMeans(returns_mat, na.rm = TRUE) * 52
  Sigma <- cov(returns_mat, use = "pairwise.complete.obs") * 52
  # Dmat and dvec for solve.QP
  Dmat <- 2 * Sigma  # QP minimizes (1/2) * w' D w, so we multiply by 2
  dvec <- rep(0, length(mu))
  
  # Constraints: sum(weights) = 1, and weights >= 0
  n_assets <- length(mu)
  # Equality constraint: sum(weights) = 1
  A_eq <- matrix(1, nrow = n_assets, ncol = 1)
  # Inequality constraints: weights >= 0
  A_ineq <- diag(n_assets)
  # Combine constraint matrix
  Amat <- cbind(A_eq, A_ineq)  # Each column is a constraint
  # Right-hand side of constraints
  bvec <- c(1, rep(0, n_assets))  # sum(weights) = 1 and weights >= 0
  # Solve QP
  sol <- solve.QP(Dmat, dvec, Amat, bvec, meq = 1)
  
  weights <- sol$solution
  
  port_ret <- as.numeric(t(weights) %*% mu)
  port_var <- as.numeric(t(weights) %*% Sigma %*% weights)
  port_sd <- sqrt(pmax(port_var, 0))
  port_sharpe <- port_ret / port_sd

  mean_var_df <- data.frame(
    ETF = colnames(returns_mat),
    Weights = round(as.numeric(weights), 6)
  )
  
  Port_Ret[regime] <- port_ret
  Port_Sd[regime] <- port_sd
  Port_Sharpe[regime] <- port_sharpe
  portfolio_list[[regime]] <- mean_var_df
  
  }

  port_per_df <- data.frame(
    Regime = names(Port_Ret),
    Portfolio_Return = Port_Ret,
    Portfolio_SD = Port_Sd,
    Portfolio_Sharpe = Port_Sharpe
  )


  return(list(
    performance = performance_list,
    correlation = correlation_list,
    portfolio =  portfolio_list,
    portfolio_perf = port_per_df
  ))
}

```


```{r,results = 'asis'}

portfolio_results <- summarise_regime(ff_returns_with_regimes, Regimes)

for (regime in Regimes) {
  
  output <- portfolio_results$performance[[regime]]
  rownames(output) <- NULL
  
  cat("\\newpage\n\n")
  
  print(knitr::kable(
    output,
    digits = 3,
    caption = paste("FF Double Sort Portfolios Perf. in", regime, "regime")
  ))
  # Page break after each table (except the last)
    if(regime != tail(Regimes, 1)) {
    cat("\\newpage\n\n")
    }
}
```

#################################################################################
# Global Factor Data Base
#################################################################################

## Data Download 

```{r}
# Create Monthly Returns 
```

## Data Merging with Global Macro Data

```{r}
# Example from the Fama French Factors 

ff_returns_with_regimes <- ff_ports_long %>%
  left_join(
    macro_data_df %>%
      dplyr::select(date, macro_regime,combined_regime) %>%
      rename(month_date = date),
    by = "month_date") %>%
  dplyr::select(-month_date)


ff_returns_with_regimes <- ff_returns_with_regimes %>%
  filter(date >= as.Date("1990-02-01")) %>%
  filter(!is.na(macro_regime))
 

knitr:: kable(ff_returns_with_regimes %>%
  group_by(macro_regime) %>%
  summarise(n_obs = sum(!is.na(return))))
```

### Analysis

```{r, results = 'asis'}
# Summarise_regime is a function introduced earlier in this document 

portfolio_results <- summarise_regime(ff_returns_with_regimes, Regimes)

for (regime in Regimes) {
  
  output <- portfolio_results$performance[[regime]]
  rownames(output) <- NULL
  
  cat("\\newpage\n\n")
  
  print(knitr::kable(
    output,
    digits = 3,
    caption = paste("FF Double Sort Portfolios Perf. in", regime, "regime")
  ))
  # Page break after each table (except the last)
    if(regime != tail(Regimes, 1)) {
    cat("\\newpage\n\n")
    }
}
```

## Outputs

### Plot Correlations

```{r, fig.width=8, fig.height=8}
for (regime in Regimes) {
  corr_matrix <- results$correlation[[regime]]
  
  corrplot(
    corr_matrix,
    method = "color",
    type = "upper",
    order = "hclust",
    addCoef.col = "black",
    tl.cex = 0.7,         # Smaller label size
    number.cex = 0.6,     # Smaller number size
    col = colorRampPalette(c("blue", "white", "red"))(200),
    cl.pos = "n",
    title = paste("Correlation –", regime),
     mar = c(1, 1, 3, 1),
  )
}
```


### Plot Portfolio Weights

```{r}
portfolio_list_renamed <- imap(results$portfolio, ~ rename(.x, !!.y := Weights))
portfolio_merged <- reduce(portfolio_list_renamed, full_join, by = "ETF")

## Bar Chart
portfolio_merged_long <- pivot_longer(portfolio_merged,cols = -ETF, names_to = "Portfolio", values_to = "Weight")
ggplot(portfolio_merged_long, aes(x = ETF, y = Weight, fill = Portfolio)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(title = "Portfolio Weights by Country", y = "Weight", x = "Asset Class") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    plot.title = element_text(hjust = 0.5)
  ) +
  coord_flip()
```

### Plot Portfolio Performance

```{r}
Performance_Analysis <- results$portfolio_perf
rownames(Performance_Analysis) <- NULL
knitr::kable(cbind(Regimes,Performance_Analysis), digits = 3,
             caption = "Performance Analysis MVO Portfolios")
```
